\chapter{USA}

\section{The Constitutional Foundation}

The 10th Amendment is kind of The Root of Everything. The US Constitution (ratified 1788) does \textbf{not} mention education anywhere \cite{us-constitution}. The 10th Amendment states:

\begin{important}
	``The powers not delegated to the United States by the Constitution, nor prohibited by it to the States, are reserved to the States respectively, or to the people.'' \cite{us-constitution}
\end{important}

This one sentence is why the USA has \textbf{no national curriculum, no national textbook board, and no national textbook.} \cite{us-constitution} Education is constitutionally a \textbf{state power}, not a federal power.


\vspace{0.5cm}

\section{50 States, 50 Systems}

\vspace{0.2cm}

The USA has \textbf{50 states}. Each state legally operates its own education system \cite{us-ed-gov-k12}. Technically, there is no single ``American education system''. There are \textbf{50 systems} running in parallel, with some federal overlay (explained in Section 2). {The US Department of Education (DoED)} was created by Congress in 1979 via the Department of Education Organization Act \cite{us-doed-act}. It does \textbf{not} write curriculum, mandate textbooks, set graduation requirements, run any school. Federal Government distribute federal \textbf{funding} to states (via formulas and grants) \cite{us-ed-gov-k12}, enforce federal \textbf{civil rights laws} in schools (\eg no discrimination) \cite{us-doed-act}, collect education data and statistics nationally \cite{nces-digest}, run federal student loan programs for higher education \cite{nces-ipeds} & set conditions tied to federal money (see Section 3). Under the Trump administration (returned January 2025), the DoED has been significantly restructured \cite{us-doed-2026}. An executive order was signed targeting its dismantlement. Many functions were transferred to other departments (\eg student loans to Treasury, special education to Health and Human Services). As of mid-2026, the DoED exists in a drastically reduced form. \textbf{This is an active, ongoing situation --- not yet fully resolved by legislation.} Even when the DoED existed at full power, it never had the authority. Its dismantlement makes curriculum even more decentralized. Each of the 50 states has a State Board of Education and a State Department of Education [18]. These bodies write or adopt academic content standards, set graduation requirements, set teacher certification requirements, determine school year length, and handle state funding distribution to local districts.

\vspace{0.3cm}

Academic Standards are \textbf{not} textbooks \cite{us-ed-gov-k12}. They are documents that say, for example: ``By the end of 4th grade, students should be able to multiply two-digit numbers.'' Standards tell you \textbf{what} to teach. They do not tell you \textbf{which book} to use to teach it.

\vspace{0.3cm}
{{\large\textbf{The Common Core Experiment (2010--present)}}

In 2010, the National Governors Association and the Council of Chief State School Officers (non-government bodies) created the Common Core State Standards (CCSS) for Math and English Language Arts. These standards were voluntary—no law forced states to adopt them, though the federal government incentivized adoption via the Race to the Top grant program. At peak adoption, 45 states plus DC adopted Common Core. However, as of 2023–2024, several states withdrew (including Indiana in 2014, Oklahoma in 2014, and South Carolina in 2015, while Texas, Alaska, and Virginia never adopted them). Ultimately, Common Core is not a national curriculum, but rather a voluntary alignment attempt that partially fragmented.

Research shows: in approximately \textbf{43 states}, curriculum selection authority rests primarily with \textbf{local school districts}, not the state government \cite{textbook-adoption-states}. This means: even within one state, different school districts can teach using different textbooks and curricula.

%%─────────────────────────────────────────────────────────────────────────────


A school district is a local educational agency (LEA) — a government body that directly runs public schools in a defined geographic area. The USA has approximately 13,000 school districts (US Census Bureau data). Each district has a locally elected or appointed School Board that decides which textbooks to purchase, which curriculum programs to implement, and local school policies and tax levies. In districts with local curriculum authority, a 7th-grade math teacher in one Texas district might use a completely different textbook than a teacher in a neighboring district \cite{textbook-adoption-states}. There is no law against this.

\section{The Textbook System in America}

\textbf{Adoption States vs.\ Open Territory States}

\textbf{Adoption States ($\sim$19--20 states):} The state government maintains an \textbf{official approved list} of textbooks for K-12 subjects \cite{textbook-adoption-states}. Districts must select from that approved list. Examples: Texas, California, Florida, North Carolina, Virginia.

\textbf{Open Territory States ($\sim$30--31 states):} No state-approved list \cite{textbook-adoption-states}. Local districts choose any textbook from any publisher.


Texas ($\sim$5.4M K-12 students) and California ($\sim$6M K-12 students) are the two largest state markets \cite{nces-digest}. Because they are adoption states with large budgets, textbook publishers \textbf{design national editions to satisfy Texas and California first}. Then sell the same books to the rest of the country. Texas and California effectively shape what's in American textbooks nationwide, even though they have no legal authority over other states \cite{textbook-adoption-states}. The USA has a \textbf{private, for-profit textbook publishing industry} \cite{textbook-adoption-states}. Major publishers: Pearson, McGraw-Hill, Houghton Mifflin Harcourt (HMH), Bedford Freeman \& Worth, Cengage. These are corporations. Schools and districts \textbf{purchase} textbooks with public (tax) funds. In public K-12, students use textbooks owned by the school and return them at year end \cite{us-ed-gov-k12}. Students don't usually buy or keep them.


\section{The K-12 Grade Structure}


\begin{center}\small
	\begin{tabularx}{0.9\textwidth}{@{} l l X @{}}
		\toprule
		\textbf{Level}            & \textbf{Grades} & \textbf{Approximate Ages} \\
		\midrule
		Kindergarten              & K               & 5--6                      \\
		Elementary School         & 1--5 or 1--6    & 6--11                     \\
		Middle School/Junior High & 6--8 or 7--8    & 11--14                    \\
		High School               & 9--12           & 14--18                    \\
		\bottomrule
	\end{tabularx}
\end{center}


``K-12'' = Kindergarten through Grade 12. Approximately 13 years of schooling before university \cite{us-ed-gov-k12}.


The exact split varies by state and district \cite{us-ed-gov-k12}. Some places run K-8 then 9-12 high school. There is no single mandated structure nationally. Each state may have its own \textbf{state assessment tests} (STAAR in Texas, MCAS in Massachusetts, PARCC in some states). These vary enormously. Historically, students took the \textbf{SAT} (College Board) or \textbf{ACT} (ACT Inc.) for college admissions --- but these are \textbf{private standardized tests}, not government exams, and many universities have now made them optional \cite{sat-act}.


\section{Structure of Higher Education}


\begin{center}\small
	\begin{tabularx}{0.97\textwidth}{@{} >{\raggedright\arraybackslash}p{3.5cm} X @{}}
		\toprule
		\textbf{Type}         & \textbf{Description}                                                  \\
		\midrule
		Community College     & 2-year public colleges, offer Associate degrees, low cost             \\
		Liberal Arts College  & 4-year private colleges, small, focus on undergraduate education      \\
		State University      & Public 4-year universities, funded by state government                \\
		Research University   & Public or private, offer Bachelor + Master + PhD, major research      \\
		For-Profit University & Private companies operating universities (controversial, many closed) \\
		\bottomrule
	\end{tabularx}
\end{center}


The USA has approximately \textbf{5,000+ degree-granting institutions} \cite{nces-ipeds}. Each American university grants its \textbf{own degree} in its own name. No central body grants the actual degree. Each university runs its own \textbf{individual admissions process} evaluating: high school GPA and transcript, SAT/ACT scores (if required), essays, extracurriculars, and letters of recommendation. Thousands of universities = thousands of separate applications and processes. Universities choose their own degree programs, textbooks (professors choose individually), and assessment methods \cite{nces-ipeds}. A professor teaching Introductory Statistics at University of Michigan can assign a completely different textbook than a professor at UCLA. \textbf{University textbooks are expensive private market products.} A single textbook can cost \$100--\$300+ USD \cite{textbook-cost-crisis}. This is a well-documented crisis in American higher education.

\section{Accreditation}

Since the government doesn't run universities, quality is controlled through \textbf{accreditation} --- a peer-review process conducted by private non-profit bodies \cite{accreditation-chea}.

\section{The Seven Regional Accrediting Bodies}


\begin{center}\small
	\begin{tabularx}{0.85\textwidth}{@{} l X @{}}
		\toprule
		\textbf{Body} & \textbf{Region}               \\
		\midrule
		NECHE         & New England                   \\
		MSCHE         & Middle States                 \\
		HLC           & North Central (large area)    \\
		SACSCOC       & Southern states               \\
		WSCUC         & Western (CA, HI, Pacific)     \\
		NWCCU         & Northwest                     \\
		ACCJC         & California community colleges \\
		\bottomrule
	\end{tabularx}
\end{center}


A university \textbf{must} be regionally accredited for its degrees to be widely recognized and for students to use federal financial aid \cite{accreditation-chea}. Unaccredited institutions' degrees are generally not recognized by employers or other universities. These bodies are \textbf{private non-profits}, not government agencies \cite{accreditation-chea}. The government recognizes them but does not operate them. This is a fundamentally American model --- private self-regulation with government recognition.


%%─────────────────────────────────────────────────────────────────────────────
\clearpage
\partbanner{PART TWO: EUROPEAN COUNTRIES}


%%─────────────────────────────────────────────────────────────────────────────

\chapter{EHEA}

\section{What Is It?}

The \textbf{Bologna Process} is an intergovernmental agreement (not an EU initiative) that began with the Bologna Declaration signed in 1999 by 29 European education ministers \cite{bologna-process}.

Goal: Make higher education degrees \textbf{comparable and compatible} across Europe, so a degree from Poland is recognizable in Spain, and students can transfer credits across borders \cite{bologna-process}.

\section{The European Higher Education Area (EHEA)}

The EHEA is the zone created by the Bologna Process.
\begin{itemize}
	\item \textbf{49 countries} are members (as of 2024): all 27 EU members + non-EU countries like UK, Switzerland, Turkey, Serbia, Ukraine, \etc (Russia suspended 2022) \cite{bologna-process}.
	\item This is \textbf{larger than the EU}.
\end{itemize}

\section{The Three-Cycle Degree System}


\begin{center}\small
	\begin{tabularx}{0.85\textwidth}{@{} l l X @{}}
		\toprule
		\textbf{Cycle} & \textbf{Degree} & \textbf{Typical Duration}  \\
		\midrule
		First          & Bachelor's      & 3 years (180 ECTS credits) \\
		Second         & Master's        & 2 years (120 ECTS credits) \\
		Third          & Doctorate/PhD   & 3+ years                   \\
		\bottomrule
	\end{tabularx}
\end{center}

\textbf{ECTS} = European Credit Transfer and Accumulation System. 1 ECTS $\approx$ 25--30 hours of student workload \cite{bologna-process}.


\section{What Bologna Does NOT Do}

Bologna does \textbf{not} harmonize: course content or curriculum, textbooks, admission standards, grading systems, or tuition fees \cite{bologna-process}. Each country still controls its own university content. Bologna only standardizes the \textbf{structure and duration} of degrees, and enables credit transfer.


%%─────────────────────────────────────────────────────────────────────────────

\chapter{Finland}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{finland-national-curriculum}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{finland-national-curriculum}.
	\item Teachers have high autonomy in choosing teaching materials \cite{finland-national-curriculum}.
	\item No high-stakes testing during K-9 \cite{finland-national-curriculum}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}


\vspace{1em}
\noindent\textcolor{Rule}{\rule{\linewidth}{0.4pt}}
\vspace{0.5em}




%%─────────────────────────────────────────────────────────────────────────────

\chapter{Norway}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{norway-udir}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{norway-udir}.
	\item Teachers have high autonomy in choosing teaching materials \cite{norway-udir}.
	\item No high-stakes testing during K-9 \cite{norway-udir}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────



\chapter{South Korea}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{korea-moe}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{korea-moe}.
	\item Teachers have high autonomy in choosing teaching materials \cite{korea-moe}.
	\item No high-stakes testing during K-9 \cite{korea-moe}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────



\chapter{Japan}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{japan-mext}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{japan-mext}.
	\item Teachers have high autonomy in choosing teaching materials \cite{japan-mext}.
	\item No high-stakes testing during K-9 \cite{japan-mext}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────


\chapter{Singapore}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{singapore-moe}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{singapore-moe}.
	\item Teachers have high autonomy in choosing teaching materials \cite{singapore-moe}.
	\item No high-stakes testing during K-9 \cite{singapore-moe}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────


\chapter{Netherlands}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{netherlands-oi}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{netherlands-oi}.
	\item Teachers have high autonomy in choosing teaching materials \cite{netherlands-oi}.
	\item No high-stakes testing during K-9 \cite{netherlands-oi}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────

\chapter{Switzerland}


\section{Basic structure}

Finland's education is controlled at the national level through the Finnish National Agency for Education (EDUFI), which writes the \textbf{National Core Curriculum} \cite{swiss-sbfi}. Local municipalities and schools then develop their own curricula based on this framework.

Key facts:
\begin{itemize}
	\item No national standardized tests until the Matriculation Examination (\textit{Ylioppilastutkinto}) at end of upper secondary \cite{swiss-sbfi}.
	\item Teachers have high autonomy in choosing teaching materials \cite{swiss-sbfi}.
	\item No high-stakes testing during K-9 \cite{swiss-sbfi}.
	\item Consistently high PISA rankings (OECD) \cite{finland-pisa}.
\end{itemize}



%%─────────────────────────────────────────────────────────────────────────────


